### 2023-06-22 tex/examples.tex
[ writing ] expand and reorganise examples

In this section, we provide three example uses of semantic traversals: generic
renaming and substitution, a usage elaborator, and a denotational semantics.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
Renaming and substitution are essential results, while the latter two examples
focus on usage annotations.

A result we will use throughout this section is \emph{reification}.
When we have an index-preserving mapping from usage-checked variables to
$\V$-environments, we can construct environments of the form
$\Gamma \env\V \Gamma$ (identity environments) for all $\Gamma$.
This lets us write the \AgdaFunction{reify} function, which  simplifies our
obligations in giving a \AgdaRecord{Semantics} by coercing
\AgdaFunction{Kripke} functions into just
\AgdaBound{$\C$}-values in an extended context.

%To show this, we instantiate the Kripke function with the renaming
%$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
%pass it an argument environment
%$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
%extended part.
%Post-composition of a renaming onto an environment comes from
%\cref{thm:env-postren}.

%\ExecuteMetaData[\Syntactictex]{reify}

\begin{lemma}[\AgdaFunction{reify}]\label{thm:reify}
  If $\V$ is an open family such that there is a function
  $v : \forallb{{\sqni} \dotto \V}$, we get a function of type
  $\forallb{\mathrm{Kripke}\,\V\,\C \dotto \mathrm{Scope}\,\C}$ for any $\C$.
\end{lemma}
\begin{proof}
  Let $b : \mathrm{Kripke}\,\V\,\C\,\Delta\,\Gamma\,A$.
  That is, $b$ is a Kripke function yielding $\C$-computations
  We want to apply $b$ so as to get a $\C\,\plr{\Gamma, \Delta}\,A$.
  Let $\grP\gamma = \Gamma$ and $\grQ\delta = \Delta$.
  The $\Box^r$ in the type of $b$ allows us to reverse-rename $\Gamma$ to
  $\Gamma, \gr0\delta$.
  Then we give the $\wand$-function an argument in context
  $\gr0\gamma, \Delta$, noting that
  $\plr{\Gamma, \gr0\delta} + \plr{\gr0\gamma, \Delta} = \plr{\Gamma, \Delta}$,
  as we wanted from the result.
  The argument needs type $\gr0\gamma, \Delta \env\V \Delta$.
  We produce this via \cref{thm:env-postren} from an environment
  $\rho : \gr0\gamma, \Delta \env\V \gr0\gamma, \Delta$ created using $v$
  and a renaming which is the complement to that used on $\Box^r$.
\end{proof}

All of the $\V$s used in examples in this paper support identity environments.
However, \citet[p.~27]{AACMM21} give some important examples that do not
support identity environments, and thus cannot use \AgdaFunction{reify}
(\cref{thm:reify}).
The feature that causes the lack of support for identity environments is that
a semantics can make use of the fact that only variables of particular kinds
are bound by the syntax.
In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
binds variables that synthesise their type, as opposed to those whose type is
checked.
The semantics of type-checking and elaboration rely on variables synthesising
their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
only binds such synthesising variables.
Similar phenomena would appear in, say, a call-by-value language where
variables are values (not computations), or a polarised language where
variables always have a polarity matching their type.

\subsection{Renaming and substitution}\label{sec:kit-to-sem}

In an unpublished note, \citet{McBride05} gives a parametrised traversal
yielding homomorphisms of syntax, the canonical examples of which are
simultaneous renaming and simultaneous substitution.
The parameters are collected in the record \AgdaRecord{Kit}.
We make a minor change to the original presentation, where instead of our
\AgdaField{ren\textasciicircum{}$\V$} field, \citeauthor{McBride05} has the
field \AgdaField{wk} allowing only context extensions.
As for the other two fields, \AgdaField{vr} allows us to map variables to
$\V$-values, so as to put newly bound variables in environments; and
\AgdaField{tm} allows us to extract terms from $\V$-values, as required when
we use the environment to handle a free variable.

\ExecuteMetaData[\Syntactictex]{Kit}

Where \citeauthor{McBride05} gave the traversal explicitly, we go via our
generic semantic traversal.
The first two fields of \AgdaRecord{Semantics} derive directly from fields of
\AgdaRecord{Kit}.
Meanwhile, to handle term constructors, we first \AgdaFunction{reify} to get a
collection of traversed subterms, and then use \AgdaInductiveConstructor{`con}
to assemble these subterms into a similarly shaped syntactic form as we started
with.
The \AgdaField{vr} field is used implicitly in \AgdaFunction{reify}, as it is
used to show that $\V$-identity environments exist.

\ExecuteMetaData[\Syntactictex]{kit-to-sem}

The action of a syntactic traversal on logical rules is basically fixed: we
preserve the logical rule and extend the environment with any newly bound
variables according to \AgdaField{vr}.
Meanwhile, the action on variables is relatively unconstrained: we look up the
variable in the environment to get a $\V$-value, then transform that $\V$-value
into a term using \AgdaField{tm}.

The idea of simultaneous renaming is that variables replace variables, whereas
with simultaneous substitution, terms replace variables.
This translates to environments for renaming containing $\sqni$-values
(variables), and environments for substitution containing $\vdash$-values
(terms).

%To implement renaming and substitution for terms, we now just implement
%syntactic kits for variables and terms, respectively.

\ExecuteMetaData[\Syntactictex]{Ren-Kit}

Notice that \AgdaFunction{ren\textasciicircum$\vdash$}, witnessing the fact
that terms are renameable, is a corollary of \AgdaFunction{Ren-Kit}.

\ExecuteMetaData[\Syntactictex]{Sub-Kit}

\subsection{A usage elaborator}\label{sec:usage-elaborator}

Using the constructs we have seen so far, producing example terms soon becomes
extremely tedious.
We can achieve some abbreviation by using pattern synonyms to wrap around
\AgdaInductiveConstructor{`con} expressions, but we still have to
produce essentially bespoke proofs whenever we use a usage-sensitive part of the
syntax.
The size of each of these proofs is roughly proportional to the number of free
variables, so the amount of proof we have to write grows roughly quadratically
with the size of terms.
An additional factor, which we can't see on paper, is that type checking time
for these proofs soon becomes prohibitive to interactive development.

Our aim in this subsection is to automate usage constraint proofs, making terms
both easier to write and more performant to check.
We invoke the automation by writing terms in a syntax where usage constraints
have been trivialised, and then use a semantic traversal over the simplified
syntax to try to produce a fully elaborated term in the original syntax.
We write the automation in a way that is generic in the syntax description, thus
avoiding repetition and facilitating the prototyping of new type systems.

@@

\subsection{A denotational semantics}

@@

>>>>

In this chapter, I provide three example uses of semantic traversals: generic
renaming and substitution\todo{Moved}, a usage elaborator, and a denotational semantics.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
Renaming and substitution are essential results, while the latter two examples
focus on usage annotations.

\section{A usage elaborator}\label{sec:usage-elaborator}

Using the constructs we have seen so far, producing example terms soon becomes
extremely tedious.
We can achieve some abbreviation by using pattern synonyms to wrap around
\AgdaInductiveConstructor{`con} expressions, but we still have to
produce essentially bespoke proofs whenever we use a usage-sensitive part of the
syntax.
The size of each of these proofs is roughly proportional to the number of free
variables, so the amount of proof we have to write grows roughly quadratically
with the size of terms.
An additional factor, which we can't see on paper but is nonetheless
significant, is that type checking time
for these proofs soon becomes prohibitive to interactive development.

Our aim in this section is to automate usage constraint proofs, making terms
both easier to write and more performant to check.
We invoke the automation by writing terms in a syntax where usage constraints
have been trivialised, and then use a semantic traversal over the simplified
syntax to try to produce a fully elaborated term in the original syntax.
We write the automation in a way that is generic in the syntax description, thus
avoiding repetition and facilitating the prototyping of new type systems.

@@

\section{A denotational semantics}\label{sec:den-sem}

@@

\section{Translating between $\name$ and L/nL}\label{sec:lnl}
\input{lnl.tex}

### 2023-06-22 tex/lnl.tex
[ writing ] expand and reorganise examples

We can express Benton's linear/non-linear logic~\cite{Benton94} in the
framework.
There are two apparent discrepancies between L/nL and what we have seen so far
to be expressible in the framework: the existence of multiple judgement modes
and the restriction on the kinds of assumptions based on the mode.
We will start by addressing these discrepancies, before presenting the encoding
in full and proving the resulting system logically equivalent to
$\lambda\gr{\mathcal R}$.

\subsection{Encoding L/nL}

In L/nL, we have a \emph{linear} and an \emph{intuitionistic} mode, with
separate types (respectively $A$ and $X$) and typing judgements (respectively
$\vdash_{\mathcal L}$ and $\vdash_{\mathcal C}$) connected by modalities $F$ and
$G$.
The mode of the conclusion type is the same as the mode of the judgement, so
only a distinction in the conclusion mode is necessary.
To achieve this distinction in the types, we have an indexed type family
$\mathrm{Ty} : \mathrm{Frag} \to \mathrm{Set}$, and set the framework types to
be the type $(f : \mathrm{Frag}) \times \mathrm{Ty}\,f$.
% \AgdaDatatype{Ty}\AgdaSymbol{:}\AgdaDatatype{Frag}\AgdaSymbol{$\to$}
% \AgdaDatatype{Set}

In L/nL, while linear judgements can have both linear and intuitionistic
assumptions, intuitionistic judgements can only have intuitionistic assumptions.
We encode this invariant on intuitionistic judgements using the $\Box$
premise combinator on all intuitionistic conclusions and wherever
intuitionistic premises occur for linear conclusions.

@@

The translations largely follow Benton's originals.

>>>>

In \cref{sec:dup-lnl}, I gave what I claimed to be an encoding of
Linear/non-Linear logic~\citep{Benton94} as a syntax description.
In this section, I rigorously state and prove the correspondence between
\citeauthor{Benton94}'s definition of L/nL and my encoding of it.
Then, I give translations from this encoding to my encoding of $\name$, and
vice versa, using two generic semantic traversals.
These results together should give us confidence that the encoding of L/nL is
correct up to logical equivalence.

\subsection{Encoding L/nL}\label{sec:encoding-LnL}

I will present translations between the systems given by
\cref{fig:LnL-orig,fig:LnL-bunched}.

I use $S$ to range over both linear and intuitionistic variables.
In this section, I use the notations $\Gamma \vdash A$ and $\Gamma \vdash X$
without subscripts on the turnstile to refer to the encoded version of the
calculus.
This notation keeps the encoded calculus distinct from the reference L/nL
calculus I am translating from and to.

The main difference between the original L/nL calculus and the encoded version
is that the encoded version contains some extra ``junk'', not corresponding to
anything in the original L/nL calculus.
This junk includes all of the wrinkles we saw when translating to and from DILL
in \cref{sec:trans-dill} --- where in the semiring-annotated system, variables
annotated $\gr\omega$ (corresponding to intuitionistic variables) can slip into
having annotation $\gr1$ or $\gr0$ whenever there are any algebraic
manipulations of the context.
In linear/non-linear logic, this slipping causes extra problems because
intuitionistic variables are supposed to be of a distinct sort to other
(i.e., linear) variables.
Additionally, we have no means in the framework to correlate types with usage
annotations, so we must deal with free variables carefully to ensure the
required correlation between linear and intuitionistic types and annotations.

With the above remarks in mind, I take it as clear how to translate the original
L/nL calculus into the encoded version, and just state the type of the
translation in \cref{thm:lnl-to-enc} without including a proof.
In contrast, I spend most of this subsection on the reverse translation, which
I provide a proof sketch of in \cref{thm:enc-to-lnl}.

\begin{proposition}\label{thm:lnl-to-enc}
  We can construct the following translations.
  \begin{align}
    (\Theta \vdashC X) &\to (\gr\omega\Theta \vdash X) \\
    (\Theta; \Gamma \vdashL A) &\to (\gr\omega\Theta, \gr1\Gamma \vdash A)
  \end{align}
\end{proposition}

The key property needed to sensibly do the reverse translation is
\emph{linear well-formedness}, as given by \cref{def:lin-well-formed}.
Linear well-formedness says that variables of linear type have linear usage
annotations.
It does not say anything about intuitionistic types and usage annotations for
two reasons.
First, talking about $\gr\omega$ is not sufficiently stable.
As we work up a derivation, the ``slip'' described earlier says that usage
annotations will tend to get larger, i.e.\ more precise.
Therefore, it makes more sense to make conditions of being greater than or equal
to some collection of usage annotations.
Second, it turns out to be unnecessary to add any conditions on variables with
intuitionistic type, because we can just treat them as if they were annotated
$\gr\omega$.
We can forget the specificness of annotations $\gr0$ and $\gr1$ when not needed,
because whatever a specifically annotated variable can do can be done by an
$\gr\omega$-annotated variable.

\begin{definition}\label{def:lin-well-formed}
  A semiring-annotated context for L/nL is \emph{linearly well formed} when all
  linear variables are annotated either $\gr0$ or $\gr1$.
\end{definition}

\begin{lemma}\label{thm:lwf}
  If $\Gamma$ is linearly well formed and $M : \plr{\Gamma \vdash S}$, then the
  context of every subterm of $M$ is linearly well formed.
\end{lemma}
\begin{proof}
  This lemma follows by inspection of the syntax description
  (\cref{fig:LnL-bunched}).
  In the subterms, the linear variables in $\Gamma$ are only changed by binding
  new variables (all instances of which maintain linear well formedness) and by
  existing variables being shared out or coerced (which never produces
  $\gr\omega$ annotations from $\gr0$ or $\gr1$).
\end{proof}

Linear well-formedness is the only condition needed for the translation given by
\cref{thm:enc-to-lnl}.
We can translate a derivation with any such context, with no further
specification of its shape.
In particular, the shape is not calculated from an original L/nL context.

\begin{proposition}\label{thm:enc-to-lnl}
  Let $\Gamma_{\mathcal C}$ be the list of variables of intuitionistic type in
  $\Gamma$.
  Let $\Gamma_{\mathcal L}$ be the list of variables of linear type in $\Gamma$
  with usage annotation $\gr1$.
  Then, whenever $\Gamma$ is linearly well formed, we can construct the
  following translations.
  \begin{align}
    (\Gamma \vdash X) &\to \plr{\Gamma_{\mathcal C} \vdashC X} \\
    (\Gamma \vdash A) &\to
      \plr{\Gamma_{\mathcal C}; \Gamma_{\mathcal L} \vdashL A}
  \end{align}
\end{proposition}
\begin{proof}
  We proceed by mutual induction on the derivations.

  First, I consider variables.
  Suppose we have $\Gamma \sqni S$.
  If $S$ is a linear type $A$, then $\Gamma$ contains one variable of type $A$
  annotated $\gr1$, while all of the other linear variables are annotated $\gr0$
  (by linear well formedness, we have no linear variables annotated
  $\gr\omega$).
  Therefore, $\Gamma_{\mathcal L} = A$, and \TirName{$\mathcal L$-var} applies.
  Otherwise, if $S$ is an intuitionistic type $X$, then
  \TirName{$\mathcal C$-var} applies to yield $\Gamma_{\mathcal C} \vdashC X$.

  For the logical rules, linear well formedness means that all variables of
  linear type act linearly.
  Additionally, every L/nL rule requiring there to be no linear variables in
  scope is guarded by $\Boxzpt$ or $I^*$ in the syntax description, excluding
  linear variables.
  Given these behaviours, the two calculi correspond closely, and it is a matter
  of inspection (and use of \cref{thm:lwf} when using the induction hypothesis)
  to complete the proof.
\end{proof}

@@

\Citet[\S 3.3]{Benton94} gives syntactic translations back and forth between
Linear/non-Linear logic and the presentation of intuitionistic linear logic
given by \citet{BBdePH93}.
In this section, I give analogous translations between my encodings of L/nL and
$\name$ as instances of the generic traversal \AgdaFunction{semantics}.
More precisely, I instantiate $\name$ to the $\{\gr0 > \gr\omega < \gr1\}$
posemiring and restrict it to the fragment containing connectives $I$,
$\otimes$, $\multimap$, and $\oc\gr\omega$, matching the fragment of L/nL
presented by \citeauthor{Benton94} and in this section.
Notably, this fragment of $\name$ excludes $\oc\gr0$ and $\oc\gr1$.
I write $\oc\gr\omega$ as just $\oc$, as in traditional linear logic.

Under the above restrictions and conventions, \citeauthor{Benton94}'s
translations between the types of ILL and L/nL can be used verbatim,
and are reproduced in \cref{fig:lnl-lr-types}.
Notably, the image of each ILL type under the $\plr{-}^o$-translation falls in
the \emph{linear} types of L/nL.
In the other direction (the $\plr{-}^*$-translation), we make extensive use of
the $\oc$-type former to translate the intuitionistic types of L/nL.

\begin{figure}
  \centering
  \begin{subfigure}{.49\linewidth}
    \centering
    \begin{align*}
      &(-)^* : \mathrm{Ty}_\lnl \to \mathrm{Ty}_{\name} \\
      &\begin{aligned}
        I^* &= I \\
        \plr{A \otimes B}^* &= A^* \otimes B^* \\
        \plr{A \multimap B}^* &= A^* \multimap B^* \\
        \plr{FX}^* &= \oc X^* \\
        1^* &= I \\
        \plr{X \times Y}^* &= \oc X^* \otimes \oc Y^* \\
        \plr{X \to Y}^* &= \oc X^* \multimap Y^* \\
        \plr{GA}^* &= A^*
      \end{aligned}
    \end{align*}
    \begin{align*}
      &(-)^* : \mathrm{Ctx}_\lnl \to \mathrm{Ctx}_{\name} \\
      &\plr{\plr{\grR\gamma}^*}_i = \grR_i\gamma_i^*
    \end{align*}
  \end{subfigure}
  \begin{subfigure}{.49\linewidth}
    \centering
    \begin{align*}
      &(-)^\circ : \mathrm{Ty}_{\name} \to \mathrm{Ty}_{\lin} \\
      &\begin{aligned}
        I^\circ &= I \\
        \plr{A \otimes B}^\circ &= A^\circ \otimes B^\circ \\
        \plr{A \multimap B}^\circ &= A^\circ \multimap B^\circ \\
        \plr{\oc A}^\circ &= GFA^\circ
      \end{aligned}
    \end{align*}
    \begin{align*}
      &(-)^\circ : \oiw \times \mathrm{Ty}_{\name} \to \Sigma_f~\mathrm{Ty}_f \\
      &\begin{aligned}
        \plr{\gr0A}^\circ &= \plr{\lin, A^\circ} \\
        \plr{\gr1A}^\circ &= \plr{\lin, A^\circ} \\
        \plr{\gr\omega A}^\circ &= \plr{\intu, GA^\circ}
      \end{aligned}
    \end{align*}
    \begin{align*}
      &(-)^\circ : \mathrm{Ctx}_{\name} \to \mathrm{Ctx}_\lnl \\
      &\plr{\plr{\grR\gamma}^\circ}_i = \grR_i\plr{\grR_i\gamma_i}^\circ
    \end{align*}
  \end{subfigure}
  \caption{Translation of types between L/nL and $\name$}
  \label{fig:lnl-lr-types}
\end{figure}

I extend $\plr{-}^*$ to contexts pointwise on the type context.
Note that this differs from \citeauthor{Benton94}'s translation of contexts in
that the intuitionistic variables of $\lnl$ are translated using usage
annotation $\gr\omega$, rather than type former $\oc$.
A translation from $\lnl$ to DILL would probably similarly use DILL's
intuitionistic variables rather than $\oc$.

For $\plr{-}^\circ$, I use an extra step to avoid producing contexts
that are not linearly well formed per \cref{def:lin-well-formed}.
Specifically, wherever there is an annotation $\gr\omega$ in a $\name$ context,
the corresponding type is wrapped in a $G$ to make it intuitionistic.
For example, $\plr{\gr0A, \gr1B, \gr\omega C}^\circ =
\plr{\gr0A^\circ, \gr1B^\circ, \gr\omega GC^\circ}$.

In Agda code, I define the $\plr{-}^\circ$ operator on
$\oiw \times \mathrm{Ty}_{\name}$ (the one that introduces a $G$ for each
$\gr\omega$) via a \emph{view} \AgdaDatatype{LIView}~\citep{MM04}.
I define usage annotations $\gr0$ and $\gr1$ (\AgdaInductiveConstructor{u0} and
\AgdaInductiveConstructor{u1} in Agda) to be \emph{linear}, with only
$\gr\omega$ (\AgdaInductiveConstructor{u$\upomega$}) being
\emph{intuitionistic}.
\AgdaDatatype{LIView} is a view because of the existence of
\ExecuteMetaData[\LinIntViewtex]{liview-type}, and well behaved in the sense
that any two views of the same usage annotation are equal (witnessed by
\AgdaFunction{liview-prop}, not shown here).
The translation from $\name$ to $\lnl$ takes cases between linear and non-linear
annotations at many points, so having these cases expressed as a view avoids
duplication of arguments between the cases for $\gr0$ and $\gr1$.

\ExecuteMetaData[\LinIntViewtex]{Linear}
\ExecuteMetaData[\LinIntViewtex]{LIView}

\begin{theorem}\label{thm:lnl-to-lr}
  Let $S$ be an $\lnl$ type, either linear or intuitionistic.
  Then, we can translate any L/nL term to a $\name$ term as follows.
  \begin{align}
    (\Gamma \vdash_\lnl S) &\to (\Gamma^* \vdash_{\name} S^*)
  \end{align}
\end{theorem}
\begin{proof}
  The proof is mostly straightforward, largely following Benton's translation.
  Similarly to the denotational semantics of \cref{sec:den-sem}, we may use a
  \AgdaRecord{Semantics} with $\V$ set to $\sqni_{\lnl}$, and then set
  $\C\,\Gamma\,S \coloneqq \Gamma^* \vdash_{\name} S^*$.
  Whenever we have induction hypotheses of \AgdaFunction{Kripke} type, we use
  the \AgdaFunction{reify} function for $\name$ to get $\name$ terms.
  Therefore, we are essentially just doing a proof by induction on the structure
  of the input term.

  Translating the \TirName{$F$i} rule into \TirName{$\oc$i} relies
  on the equivalence between duplicability (as expressed by the $\Box$ premise
  connective) and having been scaled by $\gr\omega$ (as expressed by the
  $\gr\omega \cdot {}$ premise connective).
  This holds of the $\oiw$ semiring, but not of general semirings (and is not
  even stateable for general semirings because of the mention of $\gr\omega$).
  The same reasoning is needed when translating the introduction rules for
  intuitionistic connectives, because they always have $\Box$ed premises and
  are translated using $\oc$.

  As an example of translating an intuitionistic elimination rule, let us
  consider the \TirName{$\times$e$_0$} rule.
  I reproduce it here with explicit contexts.
  \[
    \ebrule{%
      \hypo{\Gamma' \leq \Gamma}
      \hypo{\Gamma\text{ duplicable}}
      \hypo{\Gamma \vdash_{\lnl} X \times Y}
      \infer3[$\times$e$_0$]{\Gamma' \vdash_{\lnl} X}
    }
  \]

  Recall that $\plr{X \times Y}^* = \oc X^* \otimes \oc Y^*$.
  This means that we must pattern-match the hypothesis to get variables
  $\gr\omega X^*, \gr\omega Y^*$ so as to be able to use the $X^*$ for the
  conclusion and discard the $Y^*$.
  The formal derivation is as follows.
  We are able to copy $\Gamma^*$ between all of these subterms because its usage
  annotations are the same as those of $\Gamma$, which is duplicable by the
  assumption in \TirName{$\times$e$_0$}.
  The distinction between $\Gamma'$ and $\Gamma$ is minor.
  When we apply the \TirName{$\otimes$e} rule, we use the fact that
  $\Gamma' \leq \Gamma + \Gamma$, by the fact that $\Gamma' \leq \Gamma$ and
  $\Gamma \leq \Gamma + \Gamma$.
  From then on, we need only think about $\Gamma$, which behaves well because it
  is duplicable.

  \begin{align*}
    &\ebrule{%
      \hypo{\text{IH}}
      \infer[no rule]1{\Gamma^* \vdash_{\name} \oc X^* \otimes \oc Y^*}
      \infer0[Var]{\Gamma^*, \gr1\oc X^*, \gr0\oc Y^* \vdash_{\name} \oc X^*}
      \hypo{\nabla}
      \infer2[$\oc$e]{\Gamma^*, \gr1\oc X^*, \gr1\oc Y^* \vdash_{\name} X^*}
      \infer2[$\otimes$e]{\Gamma^* \vdash_{\name} X^*}
    }
    \\
    &\text{where }\nabla \coloneqq
    \\
    &\ebrule{%
      \infer0[Var]{\Gamma^*, \gr0\oc X^*, \gr1\oc Y^*, \gr\omega X^* \vdash_{\name} \oc Y^*}
      \infer0[Var]{\Gamma^*, \gr0\oc X^*, \gr0\oc Y^*, \gr\omega X^*, \gr\omega Y^* \vdash_{\name} X^*}
      \infer2[$\oc$e]{\Gamma'^*, \gr0\oc X^*, \gr1\oc Y^*, \gr\omega X^* \vdash_{\name} X^*}
    }
  \end{align*}

  In the Agda proof, renaming is required to perform lots of minor functions
  that we would ignore on paper.
  For example, the equation $\plr{\Gamma, \Delta}^* = \Gamma^*, \Delta^*$ ---
  required when induction hypotheses have newly bound variables --- is not true
  definitionally.
  Furthermore, because of the functional representation I use for contexts, it
  is not even provable without function extensionality.
  Therefore, renaming is the simplest way to get the required coercion.
  Such renamings perhaps could be avoided in most cases if contexts were
  represented as non-functional lists.
\end{proof}

\begin{theorem}\label{thm:lr-to-lnl}
  We can translate from $\name$ to the linear fragment of $\lnl$.
  \begin{align}
    (\Gamma \vdash_{\name} A) \to (\Gamma^\circ \vdash_\lnl A^\circ)
  \end{align}
\end{theorem}
\begin{proof}
  We use the same simple induction scheme as in \cref{thm:lnl-to-lr}, but with
  the places of $\name$ and $\lnl$ switched.
  However, some of the rule translations are more complex, mainly caused by the
  complexity of the $\plr{-}^\circ$ translation on contexts.

  For variables, we must consider separately the cases where the variable being
  used is annotated $\gr1$ and $\gr\omega$.
  The case for a variable $\gr1A$ is straightforward, and translates directly
  to an $\lnl$ variable $\gr1A^\circ$.
  A variable $\gr\omega A$, however, is translated to $\gr\omega GA^\circ$, so
  we must eliminate the $G$ in order to get the conclusion $A^\circ$.
  The \TirName{$G$e} rule requires its context to be duplicable, which is true
  by the fact that all of the unused variables are annotated either $\gr0$ or
  $\gr\omega$ (because they are all less than or equal to $\gr0$), and the
  variable being used is annotated $\gr\omega$.

  Most logical rules are handled very similarly to each other, so I will just
  translate the \TirName{$\otimes$i} rule as an example.
  I reproduce it here with explicit contexts (split into their usage and typing
  contexts).

  \[
    \ebrule{%
      \hypo{\grR \leq \grP + \grQ}
      \hypo{\grP\gamma \vdash_{\name} A}
      \hypo{\grQ\gamma \vdash_{\name} B}
      \infer3[$\otimes$i]{\grR\gamma \vdash_{\name} A \otimes B}
    }
  \]

  We cannot do a na\"{i}ve translation to the corresponding application of the
  \TirName{$\otimes$i} rule of $\lnl$ because $\plr{\grR\gamma}^\circ$,
  $\plr{\grP\gamma}^\circ$, and $\plr{\grQ\gamma}^\circ$ may all have different
  typing contexts.
  For example, consider the following instance.
  There are two problems.
  First, our induction hypotheses give us contexts
  containing $C^\circ$, where our conclusion wants a context containing
  $GC^\circ$.
  Second, $\gr1GC^\circ$ is not linearly well formed, because an intuitionistic
  type is given a linear annotation.
  It is therefore difficult to work with such a sequent.

  \[
    \ebrule{%
      \hypo{\gr1 C \vdash_{\name} A}
      \hypo{\gr1 C \vdash_{\name} B}
      \infer2[$\otimes$i]{\gr\omega C \vdash_{\name} A \otimes B}
    }
    \quad\rightsquigarrow\quad
    \ebrule{%
      \hypo{\gr1C^\circ \vdash_{\lnl} A^\circ}
      \ellipsis{?}{\gr1 GC^\circ \vdash_{\lnl} A^\circ}
      \hypo{\gr1C^\circ \vdash_{\lnl} B^\circ}
      \ellipsis{?}{\gr1 GC^\circ \vdash_{\lnl} B^\circ}
      \infer2[$\otimes$i]{\gr\omega GC^\circ \vdash_{\lnl} A^\circ \otimes B^\circ}
    }
  \]

  To fix these issues, firstly I overwrite the context-splitting given by the
  input term to conform to the bottom-up style of \cref{def:DILL-bottom-up}.
  This means precisely that wherever $\gr\omega$ appears in the context of the
  conclusion, it will also appear in the context of all the hypotheses.
  This gives the following partial derivation.

  \[
    \ebrule{%
      \hypo{\gr1C^\circ \vdash_{\lnl} A^\circ}
      \ellipsis{?}{\gr\omega GC^\circ \vdash_{\lnl} A^\circ}
      \hypo{\gr1C^\circ \vdash_{\lnl} B^\circ}
      \ellipsis{?}{\gr\omega GC^\circ \vdash_{\lnl} B^\circ}
      \infer2[$\otimes$i]{\gr\omega GC^\circ \vdash_{\lnl} A^\circ \otimes B^\circ}
    }
  \]

  Then, I fix the discrepancy in types using substitution.
  In the running example, the substitution needed for both subterms is of the
  type $\gr\omega GC^\circ \env{\vdash_{\lnl}} \gr1C^\circ$, which amounts to a
  term of type $\gr\omega GC^\circ \vdash_{\lnl} C^\circ$, as follows.
  Note that \TirName{$G$e} is only applicable thanks to changing the usage
  annotation from $\gr1$ to $\gr\omega$ in the previous step.

  \[
    \ebrule{%
      \infer0[Var]{\gr\omega GC^\circ \vdash_{\lnl} GC^\circ}
      \infer1[$G$e]{\gr\omega GC^\circ \vdash_{\lnl} C^\circ}
    }
  \]

  More generally, we may have to produce substitutions of type
  $$\gr0A^\circ, \gr1B^\circ, \gr\omega GC^\circ, \gr\omega GD^\circ
  \env{\vdash_{\lnl}}
  \gr0A^\circ, \gr1B^\circ, \gr1C^\circ, \gr\omega GD^\circ.$$
  These can be produced pointwise, from substitutions of types
  $\gr0A^\circ \env{\vdash_{\lnl}} \gr0A^\circ$ and
  $\gr1B^\circ \env{\vdash_{\lnl}} \gr1B^\circ$ and
  $\gr\omega GC^\circ \env{\vdash_{\lnl}} \gr1C^\circ$ and
  $\gr\omega GD^\circ \env{\vdash_{\lnl}} \gr\omega GD^\circ$.
  We have just seen how to produce the third of these, and the rest are
  identity substitutions.

  The two sui generis rules to translate are the rules for the $\oc$-modality,
  with the $\oc$ of intuitionistic linear logic becoming the composite $FG$ in
  linear/non-linear logic.
  The way of handling \TirName{$\oc$i} is similar to the way of handling rules
  like \TirName{$\otimes$i}, but involving multiplication/scaling by $\gr\omega$
  rather than addition.
  We have a similar difficult instance, shown below, where an $\gr\omega$ gets
  specialised to a $\gr1$ via the algebraic operation.

  \[
    \ebrule{%
      \hypo{\gr\omega \leq \gr\omega \cdot \gr1}
      \hypo{\gr1 C \vdash_{\name} A}
      \infer2[$\oc$i]{\gr\omega C \vdash_{\name} \oc A}
    }
  \]

  Again, the solution is to fix up the operation to keep the more general
  $\gr\omega$, allowing us to apply \TirName{$F$i} and \TirName{$G$i}, and
  the same substitution as before possibly including an application of
  \TirName{$G$e} as necessary.

  Finally, the translation of \TirName{$\oc$e} is simple, but worth checking.
  I reproduce the rule below with explicit contexts.

  \[
    \ebrule{%
      \hypo{\grR \leq \grP + \grQ}
      \hypo{\grP\gamma \vdash_{\name} \oc A}
      \hypo{\grQ\gamma, \gr\omega A \vdash_{\name} B}
      \infer3[$\oc$e]{\grR\gamma \vdash_{\name} B}
    }
  \]

  The main thing to note is that the translation of the context of the
  right-hand premise, $\grQ\gamma, \gr\omega A$, is
  $\plr{\grQ\gamma}^\circ, \gr\omega GA^\circ$ --- i.e., the translation of
  contexts gives us a $G$ thanks to the $\gr\omega$ usage annotation.
  Therefore, we do not have to eliminate the $G$, because the right-hand subterm
  is expected to do it for us.
  Indeed, we have seen in previous cases many uses of \TirName{$G$e} already.

  I translate the \TirName{$\oc$e} rule as follows.
  As in the \TirName{$\otimes$i} case, I pick $\grPprime$ and $\grQprime$ to fit
  bottom-up form, and use the same substitutions as in that case to mend the
  terms arriving from the induction hypotheses.
  Then, just \TirName{$F$e} suffices.

  \[
    \ebrule{%
      \infer0{\grR \leq \grPprime + \grQprime}
      \hypo{}
      \ellipsis{}{\plr{\grPprime\gamma}^\circ \vdash_{\lnl} FGA^\circ}
      \hypo{}
      \ellipsis{}{\plr{\grQprime\gamma}^\circ, \gr\omega GA^\circ \vdash_{\lnl} B^\circ}
      \infer3[$F$e]{\plr{\grR\gamma}^\circ \vdash_{\lnl} B^\circ}
    }
  \]

  %In the Agda version of the translation, manipulations are complicated by the
  %fact that the $\plr{-}^\circ$ translation for contexts does not distribute
\end{proof}
